A model of urban evolution based on innovation diffusion

### 2020-06-22 Docs/Papers/EvolutionInnovation/paper/paper.tex
cr_paper

\title{A model of urban evolution based on innovation diffusion}
\author{Juste Raimbault$^{1,2,3}$\\
\mbox{}\\
$^1$Center for Advanced Spatial Analysis, University College London\\
$^2$UPS CNRS 3611 ISC-PIF\\
$^3$UMR CNRS 8504 G{\'e}ographie-cit{\'e}s\\
juste.raimbault@polytechnique.edu} % email of corresponding author

@@

% Review 1 
%  evaluation:	
%2: (accept)
%Promising work. For the future, it will be interesting to see how the model can reproduce observed properties of different cities.
%TODO discussion real system

@@

%The paper is well written and the results promising. It would have been very interesting to see how this would work for a model of a real city - there are an abundance of datasets and models (even cited by the author) that would not have been difficult to integrate.
%TODO apply model to real city (more discussion)

%In addition, it would be very useful to include a discussion on how modeling innovation at a microscopic scale (or in-between, eg at a level of suburbs) changes the results of the simulations.
%TODO discussion on microscopic scale

@@

%Typos:
%1. Evolutionary economics also study regional systems of innovation, how it diffuses and thee potential existence of spatial spillovers (Uyarra, 2010).
% typo - OK
%2. Finally, to existence of subsystems in which evolution can occur is natural through the spatial aspect of the model, ….
% typo - OK
%3. It is however significantly different in the way population are updated deterministically
% typo - OK
%4. Transformation processes are included as mutations, when random innovation appear in cities.
% typo - OK
%5. In our model, cities are characterized by their size in terms of population, and city size evolve following a spatial
%typo - OK
%6. Urban genomes are adapted such that the new innovation as an initial penetration rate ..
%typo - OK
%7. We choose simply to quantify the level of hierarchy at final time, estimated by fitting the rank-size law on populations using an Ordinary Least Squares on logarithms.
% typo - OK
%8. what means that stochastic noise is not an issue for interpreting indicator values … what means that these two last indicators need more statistical precision
% typo - OK
%9. we find that it always increase for larger population
% typo - OK
%10. it means that systems where i nteraction are more global will be less unequal. This is however not always the case in such urban dynamics model as
% typo - OK
%11. For low values of mutation rates, with find a piecewise linear
% typo - OK
%12. This peak is the strongest when innovation diffuse at broader range , what may mean that this regime is due to the emergence of regional ensemble of comparable competitivity and thus a higher chance of high utility innovation,
% typo - OK
%13. cumulated variance on the second and 91% on the third, confirming that even if correlation exist the behavior of
% typo - OK
%14. one may prefer in the compromise the points just before the transition which correspond to a local diffusion setting. Thus, local diffusion correspond to intermediate
%TODO typo
%15. urban genome can be used to simulated urban dynamics
% typo - OK
%16. model to real system …. and evoluting sub-systems
% typo - OK
%Other issues:
%1. You need to deal with text formatting as some texts are being cut out at the end of the line (observed on paragraph 1 on page 5).
%TODO pb cropping fig2 - OK?
%2. Use of compromise and compromize in the same article, (kind of linear compromize between utility … prefer in the compromise the …)
%TODO typo
%3. It is not clear why the growth rate is fixed to 0.005 and the cited paper (Raimbault, 2020c) does not justify the use of that value.
%TODO justify fixed growth rate
%4. Fig 1 is a bit confusing, you will need to find a better way of presenting it than this.To enhance readability, the mutation rate and adoption rate should be clearly associated with those rows and columns you mentioned. As it is, it's hard to read.
%TODO facet names
%5. The statement “This increase however reaches a plateau around half of the width of the world (dG = 0.5) which is more obvious for higher values of mutation rates (bottom raw) …” is also confusing as it is not clear which plotting it refers to as it does not appear to apply to all plottings. I don’t think ‘(bottom raw) ’ is correct either.
%TODO clarify
%Your arguments are strong and I hope my comments will help to fine-tune your article to make it more readable. Good luck with your submission.

@@

\begin{abstract}
% Abstract length should not exceed 250 words
The dynamics of urban systems can be understood from an evolutionary perspective, in some sense extending biological and cultural evolution. Models for systems of cities implementing elementary evolutionary processes remain however to be investigated. We propose here such a model for urban dynamics at the macroscopic scale, in which the diffusion of innovations between cities captures transformation processes (mutations) and transmission processes (diffusion), using two coupled spatial interaction models. Explorations of the model on synthetic systems of cities show the role of spatial interaction and innovation diffusion ranges on measures of diversity and utility, and the existence of intermediate ranges yielding an optimal utility. Multi-objective optimization shows how the model produces a compromise between utility and diversity. This model paves the way towards more elaborated formalizations of urban evolution.
\end{abstract}

@@

% (iii) Innovation diffusion - in ALife studies ; - in urban systems
%\cite{Eletreby5664} diffusion/evolving

Besides, a central concept bridging ALife and evolution with the study of urban systems is the concept of innovation diffusion. Within artificially evolved systems, understanding how innovation emerges and how it diffuses in the population is essential \citep{bedau2000open}. One aspect of open-ended evolution are mechanisms endogeneously producing novelty. In the case of urban systems, \cite{hagerstrand1968innovation} had already highlighted the role of a hierarchical diffusion of innovation between cities in their trajectories. The aforementioned evolutionary urban theory suggests that innovation cycles and their hierarchical diffusion is a possible explanation for the properties of urban scaling laws \citep{pumain2006evolutionary}. Evolutionary economics also study regional systems of innovation, how it diffuses and the potential existence of spatial spillovers \citep{uyarra2010evolutionary}. At the microscopic scale, several agent-based models of innovation diffusion highlight the role of social and geographical networks, both of which are embedded into urban systems \citep{zhang2019empirically}. Thus, the diffusion of innovation is a privileged entry into urban evolutionary processes.

@@

	\item The sizes of cities evolve according to their performance in terms of innovation, i.e. more innovative cities are more attractive. This stage is also deterministic, following  with $P_i(t) - P_i(t-1) = w_I\cdot \sum_j \frac{V_{ij}}{<V_{ij}>}$ where $w_I$ is a fixed growth rate fixed to $w_I = 0.005$ in experiments, and where the interaction potential is defined by
\begin{equation}
V_{ij}= \frac{P_{i}(t-1) \cdot P_{j}(t-1)}{(\sum_k P_k(t-1))^2} \cdot \exp{\left(-\frac{d_{ij}}{d_G} \cdot \prod_c \delta_{c,i,t}^{\phi_{c,t}}\right)}
\end{equation}
where $\phi_{c,t} = \sum_i \delta_{i,c,t}\cdot P_i(t-1) /\sum_{i,c} \delta_{i,c,t}\cdot P_{i}(t-1)$ is the macroscopic adoption level (globally more adopted innovations will have a higher attractivity); $d_G$ is the characteristic distance for this second spatial interaction model. This model for population dynamics was proposed in such a setting by \citep{raimbault2020indirect} without the additional innovation attractivity term.

@@

The urban evolution is stopped after a fixed number of steps $t_f$.

@@

Although \cite{raimbault:halshs-01880492} proposed to apply the model of \cite{favaro2011gibrat} on real systems of cities, we focus our model study here on synthetic systems of cities, in order to isolate intrinsic effects independently of geographical contingencies. Cities are located randomly in an uniform square space. City sizes follow a Zipf law with exponent $\alpha_0 = 1$, i.e. such that $P_i (0) = \frac{P_{max}}{i^{\alpha_0}}$ \citep{pumain2012theorie}. We do not modify this initial hierarchy in our experiments. Distances between cities $d_{ij}$ are computed as euclidian distances. We take $N=30$ cities, $P_{max} = 10^5$ and $t_f = 50$ in our experiments, corresponding to a regional or small national system of cities, on timescales of the order of a century.

@@

%%%%%%%%%%%%
\begin{figure*}[t]
	\centering
	\includegraphics[width=0.49\linewidth]{figures/averageDiversity-gravityDecay_color-innovationDecay_facet-mutationRate-earlyAdoptersRate_newInnovationHierarchy1_distriblog-normal.png}
	\includegraphics[width=0.49\linewidth]{figures/averageInnovation-gravityDecay_color-innovationDecay_facet-mutationRate-earlyAdoptersRate_newInnovationHierarchy1_distriblog-normal.png}
	\includegraphics[width=0.49\linewidth]{figures/averageUtility-gravityDecay_color-innovationDecay_facet-mutationRate-earlyAdoptersRate_newInnovationHierarchy1_distriblog-normal.png}
	\includegraphics[width=0.49\linewidth]{figures/finalHierarchy-gravityDecay_color-innovationDecay_facet-mutationRate-earlyAdoptersRate_newInnovationHierarchy1_distriblog-normal.png}
	\caption{Values of indicators obtained with the grid model exploration. Each plot give each indicator among diversity $D$, utility $U$, innovation $I$ and population hierarchy $\alpha_P$, at fixed parameters $\alpha_I = 1$, $\sigma_U = 1$ and for the log-normal utility law. Indicators are plotted as a function of gravity interaction range $d_G$, for varying values of innovation diffusion range $d_I$ (color). Sub-panels show varying values of mutation rate $\beta$ (rows) and of early adoption rate $r_0$ (columns).\label{fig:exploration}}
\end{figure*}
%%%%%%%%%%%%

@@

%%%%%%%%%%%%
\begin{table}
\caption{Summary of model parameters. We first describe synthetic setup parameters which are fixed during our experiments, and then parameters varied during experiments. We give associated process, indicative range and default value for each parameter.\label{tab:params}}
	\hspace{-0.5cm}
	\footnotesize
	\begin{tabular}{|l|l|l|l|l|l|}
	\hline
	Parameter & Not. & Process & Range & Def. \\ \hline
	Number of cities & $N$ & Spatial scale & $[10 ; 100]$ & $30$\\
	Initial hierarchy & $\alpha_0$ & System of cities & $[0.5 ; 2.0]$ & $1$\\
	Initial population & $P_{max}$ & System of cities & $[10^4 ; 10^7]$ & $10^5$\\
	Simulation steps & $t_f$ & Temporal scale & $[10 ; 100]$ & $50$\\\hline
	Gravity range & $d_G$ & Crossover & $[0 ; 2]$ & $1$\\
	Innovation range & $d_I$ & Crossover & $[0 ; 2]$ & $1$\\
	Innovation rate & $\beta$ & Mutation & $[0 ; 1]$ & $0.5$\\
	Innovation hierarchy & $\alpha_I$ & Mutation & $[0 ; 2]$ & $1$\\
	Innov. utility std. & $\sigma_U$ & Mutation & $\left[ 0.7;2\right]$ & $1$\\
	Penetration rate & $r_0$ & Mutation & $\left[0.1;0.9\right]$ & $0.5$\\
	Utility type & - & Mutation & \{n;ln\}& ln \\
	\hline
	\end{tabular}
\end{table}
%%%%%%%%%%%%

@@

We also study the distance between points averages relative to their standard deviation, in order to know the significance in comparing averages. This distance is defined by $\Delta_{ij} = 2\frac{\left|\mu_i - \mu_j \right|}{\sigma_i + \sigma_j}$ if $\mu_i$ are estimated central values (mean or median) and $\sigma_i$ estimated standard deviations. These distances have an average and median larger than one for all couples, and a first quartile larger than one for innovation and population hierarchy, while the first quartile is 0.5 for utility and 0.6 for diversity. The most constraining indicators in terms of statistical precision are thus utility and diversity. 
As in the case of normal distributions, a 95\% confidence interval of size $\sigma_2$ is achieved for $n\simeq 64$ runs, a number of 50 runs is satisfying for further experiments. %TODO new number of runs

@@

We then explore a grid of the parameter space consisting of 23,168 parameter points, with a finer step on spatial parameters $d_G$ and $d_I$, and with 50 model replications for each parameter point. We find that the type of distribution for utility has small effects for all indicators and across most parameter values, but particular cases are more interesting in the log-normal case, which we comment further. Similarly, the effect of innovation hierarchy $\alpha_I$ and of innovation utility standard deviation $\sigma_U$ do not induce strong qualitative differences, so we discuss model behavior obtained at $\alpha_I = 1$ and $\sigma_U = 1$.

The Fig.~\ref{fig:exploration} shows the behavior of indicators with remaining parameters varying, namely spatial interaction parameters $d_G$ and $d_I$, mutation rate $\beta$ and innovation adoption $r_0$. Regarding the behavior of diversity (top left panel of Fig.~\ref{fig:exploration}), we find that it always increases for larger population growth spatial interaction distances, meaning that broader population exchanges will induce a higher diversity in innovations, probably through the higher innovation dynamics induced. This increase however reaches a plateau around half of the width of the world ($d_G = 0.5$) which is more obvious for higher values of mutation rates (bottom raw): higher intrinsic innovation dampens the role of spatial interactions. An increase of innovation interaction distance $d_I$ on the contrary monotonously decreases diversity, consistently with the idea of local innovation niches which can be overridden by competing innovations of higher utility coming from further away. Finally, mutation rate has a strong quantitative impact on total diversity as expected, while higher early adoption rate will change diversity behavior only when mutation rates are also high.

@@

The last experiment we perform is the application of a multi-objective optimization procedure to the model. More particularly, one could want to optimize simultaneously the global utility but also the diversity to ensure a certain resilience in the urban system. We apply thus a bi-objective genetic algorithm on the model, trying to maximize simultaneously utility and diversity. The algorithm is the NSGA2 algorithm \citep{deb2002fast}, which we run for 10,000 generations with a population of 200 individuals on a Island distribution scheme. The algorithm is integrated into the OpenMOLE software, and the stochastic aspect is internally tackled by using an embedding strategy, i.e. by adding the number of repetitions as an additional objective to find compromise points between the number of run needed and their statistical accuracy.

We show in Fig.~\ref{fig:pareto} the obtained points, which are close to a Pareto front. We find two regimes, the first corresponding to the upper points in the plot for higher utilities but lower diversities, which should correspond to the intermediate optimal regimes identified before, and in which some kind of linear compromize between utility and diversity exists: increasing global utility is done at the cost of reducing diversity. Within this first regime, two very different parameter setting coexist, one with high innovation diffusion (light blue, points with higher utility), the other in dark blue with a more local innovation diffusion. This means that reducing the span of innovation diffusion will increase diversity as one could expect. A second part of the Pareto front, obtained after a sharp transition, allows increasing the diversity but at the price of a much lower utility (points on the bottom right). These points are all close to equivalent, and one may prefer in the compromise the points just before the transition which correspond to a local diffusion setting. Thus, local diffusion correspond to intermediate compromises, while broad diffusion corresponds to extremes in the Pareto front. In a nutshell, this optimization exercise is interesting both to show how the model produces compromises, but also how it could be used in practice for innovation policies applied to systems of cities.

@@

Several extensions and applications would be possible to this first model exploration. First, the innovation space in our model remained unidimensional. \citep{hidalgo2007product} show in terms of industrial production by countries that these industrial spaces are highly dimensional in terms of product produced and interdependencies between countries and types of production, increasing the path-dependency in economic trajectories. The investigation of patent data from a semantic viewpoint also shows the highly dimensional nature of technological innovation \citep{bergeaud2017classifying}. A direct extension of our model would consist in having a matrix genome instead of a vector one. Innovations would occur across several dimensions which can correspond to industrial or technological domains, but also social and cultural innovation and innovations in terms of infrastructures which condition the way systems of cities evolve. Each dimension should have particular innovation rules depending on its nature, and dependencies across dimensions could be introduced, implementing the possible emergence of technological co-evolution niches beside the spatial co-evolution niches.

Second, applying the model to real systems of cities, both in terms of initial parametrization and of empirical laws for innovation processes, would allow bringing this approach closer to possible policy applications. \cite{raimbault:halshs-01880492} and later \cite{2020arXiv200510007R} benchmarked several models of urban growth based on interactions between cities, and integrated the model of \cite{favaro2011gibrat} on which this work was based. It however only included accurate initial populations and distance matrix, but no empirical data on innovation waves which were also synthetically constructed. One important aspect of this application to real systems of cities is thus a test of the performance of this model in predicting possible trajectories for systems of cities.

Finally, our approach was rather restricted in the sense that even a broad geographical range is taken into account, a single ontological scale is included in the model, i.e. the macroscopic scale since cities are the basic agents. 
One can expect processes of innovation diffusion to behave differently at the microscopic scale or at intermediate scales of companies or innovation clusters % lit spillovers?
Similarly, a single temporal scale was included, although two dynamics of spatial interaction and innovation diffusion are effectively combined. A multi-scale approach of urban evolution would be necessary to fully account for the complexity of these systems. \cite{raimbault:halshs-02351722} introduced a multi-scalar model for population growth with upward and downward strong feedbacks between the mesoscopic scale (urban form) and the macroscopic scale (spatial interaction model). Adapting this approach in the case of urban evolution through innovation diffusion would be an interesting potential development.
The aspects to be clarified

@@

We have introduced a simple model of urban evolution integrating effectively the evolutionary processes of transmission, transformation and evolving sub-systems, through innovation diffusion and spatial interactions. Model exploration yield complex behavior while multi-objective optimization shows the potentiality for the model to produce compromises between utility and diversity in the system of cities. This work is thus a first step towards more elaborated models of urban evolution.

>>>>

\title{A model of urban evolution based on innovation diffusion}
\author{Juste Raimbault$^{1,2,3}$\\
\mbox{}\\
$^1$Center for Advanced Spatial Analysis, University College London\\
$^2$UPS CNRS 3611 ISC-PIF\\
$^3$UMR CNRS 8504 G{\'e}ographie-cit{\'e}s\medskip\\
juste.raimbault@polytechnique.edu} % email of corresponding author

@@

% Review 1 
%  evaluation:	
%2: (accept)
%Promising work. For the future, it will be interesting to see how the model can reproduce observed properties of different cities.
% discussion real system - OK

@@

%The paper is well written and the results promising. It would have been very interesting to see how this would work for a model of a real city - there are an abundance of datasets and models (even cited by the author) that would not have been difficult to integrate.
% apply model to real city (more discussion) - OK

%In addition, it would be very useful to include a discussion on how modeling innovation at a microscopic scale (or in-between, eg at a level of suburbs) changes the results of the simulations.
% discussion on microscopic scale - OK

@@

%Typos:
%1. Evolutionary economics also study regional systems of innovation, how it diffuses and thee potential existence of spatial spillovers (Uyarra, 2010).
% typo - OK
%2. Finally, to existence of subsystems in which evolution can occur is natural through the spatial aspect of the model, ….
% typo - OK
%3. It is however significantly different in the way population are updated deterministically
% typo - OK
%4. Transformation processes are included as mutations, when random innovation appear in cities.
% typo - OK
%5. In our model, cities are characterized by their size in terms of population, and city size evolve following a spatial
%typo - OK
%6. Urban genomes are adapted such that the new innovation as an initial penetration rate ..
%typo - OK
%7. We choose simply to quantify the level of hierarchy at final time, estimated by fitting the rank-size law on populations using an Ordinary Least Squares on logarithms.
% typo - OK
%8. what means that stochastic noise is not an issue for interpreting indicator values … what means that these two last indicators need more statistical precision
% typo - OK
%9. we find that it always increase for larger population
% typo - OK
%10. it means that systems where i nteraction are more global will be less unequal. This is however not always the case in such urban dynamics model as
% typo - OK
%11. For low values of mutation rates, with find a piecewise linear
% typo - OK
%12. This peak is the strongest when innovation diffuse at broader range , what may mean that this regime is due to the emergence of regional ensemble of comparable competitivity and thus a higher chance of high utility innovation,
% typo - OK
%13. cumulated variance on the second and 91% on the third, confirming that even if correlation exist the behavior of
% typo - OK
%14. one may prefer in the compromise the points just before the transition which correspond to a local diffusion setting. Thus, local diffusion correspond to intermediate
% typo - OK
%15. urban genome can be used to simulated urban dynamics
% typo - OK
%16. model to real system …. and evoluting sub-systems
% typo - OK
%Other issues:
%1. You need to deal with text formatting as some texts are being cut out at the end of the line (observed on paragraph 1 on page 5).
% pb cropping fig2 - OK?
%2. Use of compromise and compromize in the same article, (kind of linear compromize between utility … prefer in the compromise the …)
% typo - OK
%3. It is not clear why the growth rate is fixed to 0.005 and the cited paper (Raimbault, 2020c) does not justify the use of that value.
% justify fixed growth rate - OK
%4. Fig 1 is a bit confusing, you will need to find a better way of presenting it than this.To enhance readability, the mutation rate and adoption rate should be clearly associated with those rows and columns you mentioned. As it is, it's hard to read.
% facet names - OK
%5. The statement “This increase however reaches a plateau around half of the width of the world (dG = 0.5) which is more obvious for higher values of mutation rates (bottom raw) …” is also confusing as it is not clear which plotting it refers to as it does not appear to apply to all plottings. I don’t think ‘(bottom raw) ’ is correct either.
% clarify - OK
%Your arguments are strong and I hope my comments will help to fine-tune your article to make it more readable. Good luck with your submission.

@@

\begin{abstract}
% Abstract length should not exceed 250 words
The dynamics of urban systems can be understood from an evolutionary perspective, in some sense extending biological and cultural evolution. Models for systems of cities implementing elementary evolutionary processes remain however to be investigated. We propose here such a model for urban dynamics at the macroscopic scale, in which the diffusion of innovations between cities captures transformation processes (mutations) and transmission processes (diffusion), using two coupled spatial interaction models. Explorations of the model on synthetic systems of cities show the role of spatial interaction and innovation diffusion ranges on measures of diversity and utility, and the existence of intermediate ranges yielding an optimal utility. Multi-objective optimization shows how the model produces a compromize between utility and diversity. This model paves the way towards more elaborated formalizations of urban evolution.
\end{abstract}

@@

% (iii) Innovation diffusion - in ALife studies ; - in urban systems

Besides, a central concept bridging ALife and evolution with the study of urban systems is the concept of innovation diffusion. Within artificially evolved systems, understanding how innovation emerges and how it diffuses in the population is essential \citep{bedau2000open}. One aspect of open-ended evolution are mechanisms endogeneously producing novelty. The other way around, artificial societies approaches have also been pinned as a privileged entry for understanding the dynamics of innovation \citep{zenobia2009artificial}. In the case of urban systems, \cite{hagerstrand1968innovation} had already highlighted the role of a hierarchical diffusion of innovation between cities in their trajectories. The aforementioned evolutionary urban theory suggests that innovation cycles and their hierarchical diffusion is a possible explanation for the properties of urban scaling laws \citep{pumain2006evolutionary}. \cite{paulus2004co} provides empirical support for this approach by unveiling the co-evolution between economic functions of urban areas for the French systems of cities, which are closely related to the dynamics of innovation. Evolutionary economics also study regional systems of innovation, how it diffuses and the potential existence of spatial spillovers \citep{uyarra2010evolutionary}. At the microscopic scale, several agent-based models of innovation diffusion highlight the role of social and geographical networks, both of which are embedded into urban systems \citep{zhang2019empirically}. \cite{daude2004apports} proposes for example to understand the diffusion of an agricultural innovation within a geographical system using an agent-based model. Thus, the diffusion of innovation is a privileged entry into urban evolutionary processes, as it is furthermore tightly linked with other dimensions of urban systems such as the social and economic dimensions \citep{sonis1995territorial}.

@@

	\item The sizes of cities evolve according to their performance in terms of innovation, i.e. more innovative cities are more attractive. This stage is also deterministic, following  with $P_i(t) - P_i(t-1) = w_I\cdot \sum_j \frac{V_{ij}}{<V_{ij}>}$ where $w_I$ is a fixed growth rate, and where the interaction potential is defined by
\begin{equation}
V_{ij}= \frac{P_{i}(t-1) \cdot P_{j}(t-1)}{(\sum_k P_k(t-1))^2} \cdot \exp{\left(-\frac{d_{ij}}{d_G} \cdot \prod_c \delta_{c,i,t}^{\phi_{c,t}}\right)}
\end{equation}
where $\phi_{c,t} = \sum_i \delta_{i,c,t}\cdot P_i(t-1) /\sum_{i,c} \delta_{i,c,t}\cdot P_{i}(t-1)$ is the macroscopic adoption level (globally more adopted innovations will have a higher attractivity); $d_G$ is the characteristic distance for this second spatial interaction model. This model for population dynamics was proposed in such a setting by \citep{raimbault2020indirect} without the additional innovation attractivity term.

@@

The urban evolution is stopped after a fixed number of steps $t_f$. Note that other stopping criteria such as reaching a total population, a maximal population for the largest city, or a maximal number of innovations introduced, could be included for other applications of the model.

@@

Although \cite{raimbault:halshs-01880492} proposed to apply the model of \cite{favaro2011gibrat} on real systems of cities, we focus our model study here on synthetic systems of cities, in order to isolate intrinsic effects independently of geographical contingencies. Cities are located randomly in an uniform square space. City sizes follow a Zipf law with exponent $\alpha_0 = 1$, i.e. such that $P_i (0) = \frac{P_{max}}{i^{\alpha_0}}$ \citep{pumain2012theorie}. We do not modify this initial hierarchy in our experiments. Distances between cities $d_{ij}$ are computed as euclidian distances. We take $N=30$ cities, $P_{max} = 10^5$ and $t_f = 50$ in our experiments, corresponding to a regional or small national system of cities, on timescales of the order of a century. According to these approximates scales, we fix $w_I = 0.005$ in experiments which gives a maximal possible growth factor of $(1.005)^{50} = 1.28$ over the whole period, in consistence with mature urban systems having relatively slow population growth \citep{pumain1982dynamique}.

@@

%%%%%%%%%%%%
\begin{figure*}[t]
	\centering
	\includegraphics[width=0.49\linewidth]{figures/averageDiversity-gravityDecay_color-innovationDecay_facet-mutationRate-earlyAdoptersRate_newInnovationHierarchy1_utilityStd1_distriblog-normal}
	\includegraphics[width=0.49\linewidth]{figures/averageInnovation-gravityDecay_color-innovationDecay_facet-mutationRate-earlyAdoptersRate_newInnovationHierarchy1_utilityStd1_distriblog-normal.png}
	\includegraphics[width=0.49\linewidth]{figures/averageUtility-gravityDecay_color-innovationDecay_facet-mutationRate-earlyAdoptersRate_newInnovationHierarchy1_utilityStd1_distriblog-normal.png}
	\includegraphics[width=0.49\linewidth]{figures/finalHierarchy-gravityDecay_color-innovationDecay_facet-mutationRate-earlyAdoptersRate_newInnovationHierarchy1_utilityStd1_distriblog-normal.png}
	\caption{Values of indicators obtained with the grid model exploration. Each plot give each indicator among diversity $D$, utility $U$, innovation $I$ and population hierarchy $\alpha_P$, at fixed parameters $\alpha_I = 1$, $\sigma_U = 1$ and for the log-normal utility law. Indicators are plotted as a function of gravity interaction range $d_G$, for varying values of innovation diffusion range $d_I$ (color). Sub-panels show varying values of mutation rate $\beta$ (rows) and of early adoption rate $r_0$ (columns).\label{fig:exploration}}
\end{figure*}
%%%%%%%%%%%%

@@

%%%%%%%%%%%%
\begin{table}
\caption{Summary of model parameters. We first describe synthetic setup parameters which are fixed during our experiments, and then parameters varied during experiments. We give associated process, indicative range and default value for each parameter.\label{tab:params}}\medskip
	\hspace{-1cm}
	\footnotesize
	\begin{tabular}{|l|l|l|l|l|l|}
	\hline
	Parameter & Not. & Process & Range & Def. \\ \hline
	Number of cities & $N$ & Spatial scale & $[10 ; 100]$ & $30$\\
	Initial hierarchy & $\alpha_0$ & System of cities & $[0.5 ; 2.0]$ & $1$\\
	Initial population & $P_{max}$ & System of cities & $[10^4 ; 10^7]$ & $10^5$\\
	Simulation steps & $t_f$ & Temporal scale & $[10 ; 100]$ & $50$\\
	Growth rate & $w_I$ & Pop. growth & $[0.001 ; 0.01]$ & $0.005$\\\hline
	Gravity range & $d_G$ & Crossover & $[0 ; 2]$ & $1$\\
	Innovation range & $d_I$ & Crossover & $[0 ; 2]$ & $1$\\
	Innovation rate & $\beta$ & Mutation & $[0 ; 1]$ & $0.5$\\
	Innovation hierarchy & $\alpha_I$ & Mutation & $[0 ; 2]$ & $1$\\
	Innov. utility std. & $\sigma_U$ & Mutation & $\left[ 0.7;2\right]$ & $1$\\
	Penetration rate & $r_0$ & Mutation & $\left[0.1;0.9\right]$ & $0.5$\\
	Utility type & - & Mutation & \{n;ln\}& ln \\
	\hline
	\end{tabular}
\end{table}
%%%%%%%%%%%%

@@

We also study the distance between points averages relative to their standard deviation, in order to know the significance in comparing averages. This distance is defined by $\Delta_{ij} = 2\frac{\left|\mu_i - \mu_j \right|}{\sigma_i + \sigma_j}$ if $\mu_i$ are estimated central values (mean or median) and $\sigma_i$ estimated standard deviations. These distances have an average and median larger than one for all couples, and a first quartile larger than one for innovation and population hierarchy, while the first quartile is 0.5 for utility and 0.6 for diversity. The most constraining indicators in terms of statistical precision are thus utility and diversity. As in the case of normal distributions, a 95\% confidence interval of size $\sigma_2$ is achieved for $n\simeq 64$ runs, a number of 50 runs is satisfying for further experiments.

@@

We then explore a grid of the parameter space consisting of 23,168 parameter points, with a finer step on spatial parameters $d_G$ and $d_I$, and with 50 model replications for each parameter point. We find that the type of distribution for utility has small effects for all indicators and across most parameter values, but particular cases are more interesting in the log-normal case, which we comment further. Similarly, the effect of innovation hierarchy $\alpha_I$ and of innovation utility standard deviation $\sigma_U$ do not seem to induce strong qualitative differences, so we discuss model behavior obtained at $\alpha_I = 1$ and $\sigma_U = 1$.

The Fig.~\ref{fig:exploration} shows the behavior of indicators with remaining parameters varying, namely spatial interaction parameters $d_G$ and $d_I$, mutation rate $\beta$ and innovation adoption $r_0$. Regarding the behavior of diversity (top left panel of Fig.~\ref{fig:exploration}), we find that it always increases for larger population growth spatial interaction distances, meaning that broader population exchanges will induce a higher diversity in innovations, probably through the higher innovation dynamics induced. This increase is however slowed down around half of the width of the world ($d_G = 0.5$), which is more obvious with higher values of mutation rates for which a plateau is reached (bottom row, in particular the sub-panel for $r_0=0.1$ and $\beta = 0.5$): higher intrinsic innovation dampens the role of spatial interactions. An increase of innovation interaction distance $d_I$ on the contrary monotonously decreases diversity, consistently with the idea of local innovation niches which can be overridden by competing innovations of higher utility coming from further away. Finally, mutation rate has a strong quantitative impact on total diversity as expected, while higher early adoption rate will change diversity behavior only when mutation rates are also high.

@@

The last experiment we perform is the application of a multi-objective optimization procedure to the model. More particularly, one could want to optimize simultaneously the global utility but also the diversity to ensure a certain resilience in the urban system. We apply thus a bi-objective genetic algorithm on the model, trying to maximize simultaneously utility and diversity. The algorithm is the NSGA2 algorithm \citep{deb2002fast}, which we run for 50,000 generations with a population of 200 individuals and an Island distribution scheme. The algorithm is integrated into the OpenMOLE software, and the stochastic aspect is internally tackled by using an embedding strategy, i.e. by adding the number of repetitions as an additional objective to find compromize points between the number of run needed and their statistical accuracy.

We show in Fig.~\ref{fig:pareto} the obtained points, which are close to a Pareto front. We include the points with at least 20 stochastic samples. We find three regimes, the first corresponding to the upper points in the plot for higher utilities but lower diversities, which should correspond to the intermediate optimal regimes identified before, and in which some kind of linear compromize between utility and diversity exists (but with a very flat slope for utility): increasing global utility is done at the cost of reducing diversity. Within this first regime, two very different parameter setting coexist, one with high innovation diffusion (light blue, points with higher utility), the other in dark blue with a more local innovation diffusion. This means that reducing the span of innovation diffusion will increase diversity as one could expect. A second part of the Pareto front corresponds to intermediate points with lower utility but higher diversity, and is exclusively composed of local innovation parameters. The third part of the Pareto front, obtained after a sharp transition, allows increasing significantly the diversity compared to the first part, but at the price of a much lower utility (points on the bottom right). These points are all close to equivalent, and one may prefer to consider as a compromize the extreme points which correspond to a broad diffusion setting, but this last regime is also mixed with local innovation. Thus, local diffusion correspond to intermediate compromizes, while broad diffusion corresponds to extremes in the Pareto front (first and third regimes). In a nutshell, this optimization exercise is interesting both to show how the model produces compromizes, but also how it could be used in practice for innovation policies applied to systems of cities.

@@

Several extensions and applications would be possible to this first model exploration. First, the innovation space in our model remained unidimensional. \citep{hidalgo2007product} show in terms of industrial production by countries that these industrial spaces are highly dimensional in terms of product produced and interdependencies between countries and types of production, increasing the path-dependency in economic trajectories. The investigation of patent data from a semantic viewpoint also shows the highly dimensional nature of technological innovation \citep{raimbault2016investigating,bergeaud2017classifying}. A direct extension of our model would consist in having a matrix genome instead of a vector one. Innovations would occur across several dimensions which can correspond to industrial or technological domains, but also social and cultural innovation and innovations in terms of infrastructures which condition the way systems of cities evolve \citep{raimbault2018caracterisation}. Each dimension should have particular innovation rules depending on its nature, and dependencies across dimensions could be introduced, implementing the possible emergence of technological co-evolution niches beside the spatial co-evolution niches.

Second, applying the model to real systems of cities, both in terms of initial parametrization and of empirical laws for innovation processes, would allow bringing this approach closer to possible policy applications. \cite{raimbault:halshs-01880492} and later \cite{2020arXiv200510007R} benchmarked several models of urban growth based on interactions between cities, and integrated the model of \cite{favaro2011gibrat} on which this work was based. It however only included accurate initial populations and distance matrix, but no empirical data on innovation waves which were also synthetically constructed. One important aspect of this application to real systems of cities is thus a test of the performance of this model in predicting possible population trajectories for systems of cities. Several obstacles however remain for a more complete application of the innovation diffusion urban evolution model. Harmonized datasets with consistent definition of urban entities, in particular which can be compared between different urban systems and long time scales, are crucial to study the dynamics of urban systems as done for example by \cite{pumain2015multilevel}. Innovation measures are not included in such datasets, and existing innovation proxys are limited in time and space, as for example the US Patent Office database for which inventors have been localized \citep{morrison2017disambiguation}. Then, how to link innovation proxy measures with model processes, in other words how to parametrize or calibrate the model on real data, remains an open question. Finally, how various urban systems with very different histories and at different stages of development would fit model assumptions, in particular the fact that innovation waves partly drive growth rates of cities, remains to be empirically investigated \citep{pumain2009innovation}.

Finally, our approach was rather restricted in the sense that even a broad geographical range is taken into account, a single ontological scale is included in the model, i.e. the macroscopic scale since cities are the basic agents. One can expect processes of innovation diffusion to behave differently at the microscopic scale or at intermediate scales of companies or innovation clusters \cite{audretsch2004knowledge}. Other types of agents and processes may be more significant than spatial interaction at this finer scales, for example with a diffusion through social networks \citep{zhang2019empirically}. Similarly, a single temporal scale was included, although two dynamics of spatial interaction and innovation diffusion are effectively combined. A multi-scale approach of urban evolution would be necessary to fully account for the complexity of these systems. \cite{raimbault:halshs-02351722} introduced a multi-scalar model for population growth with upward and downward strong feedbacks between the mesoscopic scale (urban form) and the macroscopic scale (spatial interaction model). Adapting this approach in the case of urban evolution through innovation diffusion would be an interesting potential development.

@@

We have introduced a simple model of urban evolution integrating effectively the evolutionary processes of transmission, transformation and evolving sub-systems, through innovation diffusion and spatial interactions. Model exploration yield complex behavior while multi-objective optimization shows the potentiality for the model to produce compromizes between utility and diversity in the system of cities. This work is thus a first step towards more elaborated models of urban evolution.
